Progress Report for CSS 586 Course Project: \\ Modeling Latent Patterns to Generate Symbolic Music

### 2021-05-04 papers/progress/alex/progress-alex.tex
save progress report

\begin{document}
\settopmatter{printacmref=false}
\setcopyright{none}
\renewcommand\footnotetextcopyrightpermission[1]{}
\pagestyle{plain}
\title{Progress Report for CSS 586 Course Project: \\
Modeling Latent Patterns in Music}

@@

\begin{abstract}
This report explores recent research in symbolic music modeling with
deep learning and provides a progress report on a course project to
train a generative model of classical and popular music on the
MusicNet and Lakh MIDI datasets.
\end{abstract}

\keywords{deep learning, sequence learning, generative modeling,
  recurrent neural networks, music, MIDI}

@@

Depending on the task, machine learning models of music may be trained
on the audio signal itself, either in a time domain or a frequency
domain representation, or they may be trained on a digital symbolic
representation of music, the most common of which is MIDI (Musical
Instrument Digital Interface) notation. MIDI is an encoding of music
as streams of bytes in one or more tracks or channels, each
representing a sequence of 128 possible pitch values, along with
timing, pressure and instrument values. A music transcription model
may transcribe an audio signal as a MIDI score, which can easily be
converted into other symbolic representations such as sheet music for
human performers to read from, while a synthesizer model can convert
MIDI representations into audio signals. A generative music model can
be either trained to generate raw audio, or trained to produce a score
that must be played by either a synthesizer or by humans to produce an
audio music performance.

@@

The state of the art in music generation has a long way to go before
it can consistently generate music scores or performances that would
be enjoyable and popular to listen to. There are a number of
interesting, recent papers and projects in the area of music machine
learning, several of which focus on learning long-term dependencies in
music to produce longer clips with coherent structure.

@@

MusicVAE, part of the Magenta project, is a variational LSTM
autoencoder for MIDI that incorporates a novel hierarchical structure
using a ``conductor'' recurrent layer in its decoder model to better
capture structure at multiple levels and avoid the problem of
``posterior/mode collapse'' whereby a generative model learns to
ignore its latent code and rely on autoregression
\cite{roberts_hierarchical_2018}. This model is capable of generating
new melodies that blend two given melodies via latent space
interpolation.

Another Magenta model called Music Transformer is a generative model
that borrows its approach from the Natural Language Processing (NLP)
domain, using a self-attention network to model MIDI music as a
sequence of discrete tokens with relative positional dependencies
\cite{huang_music_2018}. This model was trained on the MAESTRO dataset
of Piano-e-competition performances \cite{hawthorne2019enabling} and
its generated piano music received favorable qualitative (Likert
scale) ratings from human listeners \cite{huang_music_2018}.

MuseGAN \cite{dong2017musegan} is an application of Generative Adversarial
Networks to polyphonic MIDI music generation, trained on four-bar phrases of a
multi-track pianoroll representation of rock songs from the Lakh Midi Dataset
\cite{raffel_learning-based_2016}.

A major advantage of working with the symbolic representation of music is that
it is of far lower dimensionality than the raw audio waveforms of a recorded
performance, which makes it less computationally expensive. However, there are
many aspects of musical performance that are not captured by a symbolic
representation, so the expressiveness of symbolic generative models is
constrained \cite{manzelli_conditioning_2018}.

@@

Jukebox by OpenAI utilizes a vector-quantized variational autoencoder
(VQ-VAE) to compress raw audio into a sequence of discrete codes and
models these sequences using autoregressive transformers to generate
music of various popular genres, including singing and lyrics
\cite{dhariwal2020jukebox}.

Audio data can also be modeled in the frequency domain through the use of
Fourier analysis. The recent MelNet model is trained on spectrograms and can
learn musical structures such as melody and harmony and variations in volume,
timbre and rhythm \cite{vasquez2019melnet}.

Prior work points out that the division between symbolic music notes
and the sounds of music is analogous to the division between symbolic
language and utterances in speech, which may inspire ideas for
combining the two approaches \cite{hawthorne2019enabling}. A paper
from Boston University describes an effort to combine the symbolic and
waveform approaches to music modeling, by training an LSTM to learn
melodic structure of different styles of music, then providing
generations from this model as conditioning inputs to a WaveNet-based
raw audio generator \cite{manzelli_conditioning_2018}.

@@

This research project will focus on the generative modeling of
symbolic music using MIDI data as inputs, because of the advantages of
symbolic music models in representing long-term structure in musical
compositions to produce generations with coherent structure and use of
repetition over long time scales.

@@

Two primary datasets will be used for this research project: MusicNet,
which is a collection of 330 freely licensed European classical music
recordings with aligned MIDI scores \cite{thickstun2017learning}, and
the Lakh MIDI Dataset, which includes a collection of 45,129 Creative
Commons licensed MIDI files of popular music songs that have been
matched and aligned to MP3 recordings, and of which we will use a
subset for model training and evaluation
\cite{raffel_learning-based_2016}. Including this second dataset may
help us to generalize the modeling approach beyond European classical
music to include other popular genres and associated instruments.

@@

Several choices must be made in how to preprocess binary MIDI files
into training examples for a neural network. There are multiple
open-source Python packages capable of reading MIDI files into NumPy
array-based representations, such as pretty\_midi
\cite{raffel_pretty_midi_2014} and Pypianoroll
\cite{dong_pypianoroll_2018}. In order to accommodate polyphonic
music, we will convert each MIDI file into a pianoroll representation,
as visualized in Figure \ref{pianoroll}, wherein each instrument track
is a sparse matrix that multi-hot encodes the velocity values for each
of 128 possible pitch levels at each timestep. Because songs are
typically at least a few minutes long and of varying length, it will
not be feasible to train with entire songs as examples, so we will
crop songs into phrases of equal numbers of measures to use as
training data.

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{first_four_bars.png}
  \caption{Pianoroll visualization of the first four measures of
    Beethoven's Serioso String Quartet}
  \label{pianoroll}
  \Description{A pianoroll representation of cello, viola and violin
    parts from a Beethoven piece.}
\end{figure}

Data augmentation is also possible and we plan to test its impact on
our results--the literature suggests augmentation via pitch shifting
the entire training example up or down by up to six semitones, and
increasing or reducing the speed by up to 10\% in order to create
additional training examples and reduce overfitting
\cite{oore_this_2018}.

\subsection{Model Fitting}

We will explore several modeling approaches to generating symbolic music:

\begin{itemize}
  \item{Sliding window sequence prediction with RNNs (LSTM/GRU)}
  \item{Sliding window sequence prediction with Transformers}
  \item{Latent space interpolation with Sequential VAEs}
  \item{Latent space interpolation with Sequential GANs}
\end{itemize}

@@

Evaluation of generative models is challenging because there is no
equivalent of an accuracy metric like what is used in supervised
learning. For autoencoder models we can measure how accurately the
model can reconstruct its own inputs, but this does not tell us the
quality of the interpolated examples. Generative models are typically
evaluated using a combination of qualitative metrics whereby human
judges rate the quality of the generated examples (essentially a
Turing test), and quantitative metrics that assess the differences in
the parametric distributions of generated and real examples. Yang and
Lerch (2020) proposes a set of metrics informed by music theory, for
probabilistically evaluating how similar the generations are to known
sample distributions of real music \cite{yang_evaluation_2020}. These
metrics include counts, ranges, histograms and transition matrices of
pitches and note lengths, then the Kullback-Leibler divergence and
overlapping area of the probability density functions are use to
compare against known reference distributions per musical genre
\cite{yang_evaluation_2020}. Due to the cost and time requirements
associated with designing a human subjects experiment, we plan to
utilize this quantitative approach to generation quality assessment.

@@

At this stage, we have decided on the modeling problem, downloaded the
MIDI datasets, and begun exploring them to learn how to interpret,
process, and visualize them. We have started writing a training data
processing pipeline that will provide batches of training examples to
Keras model layers for fitting.

>>>>

\begin{document}
\settopmatter{printacmref=false}
\setcopyright{none}
\renewcommand\footnotetextcopyrightpermission[1]{}
\pagestyle{plain}
\title{Progress Report for CSS 586 Course Project: \\
Modeling Latent Patterns to Generate Symbolic Music}

@@

\begin{abstract}
  This report explores recent research in symbolic music modeling with
  deep learning and provides a progress report on a course project to
  train and evaluate generative models of classical and popular music
  on the MusicNet and Lakh MIDI datasets.
\end{abstract}

\keywords{
  generative modeling,
  midi,
  music,
  recurrent neural networks,
  sequence learning,
  symbolic music,
  deep learning
}

@@

Depending on the task, machine learning models of music may be trained
on the audio signal itself, either in a time domain or a frequency
domain representation, or they may be trained on a digital symbolic
representation of music, the most common of which is MIDI (Musical
Instrument Digital Interface) notation. MIDI is an encoding of music
as streams of bytes in one or more tracks or channels, each
representing a sequence of 128 possible pitch values (where 0 is the
lowest and 127 is the highest), along with timing, pressure and
instrument identifier values. A music transcription model may
transcribe an audio signal as a MIDI score, which can easily be
converted into other symbolic representations such as sheet music for
human performers to read from, while a synthesizer model can convert
MIDI representations into audio signals. A generative music model can
be either trained to generate raw audio, or trained to produce a
symbolic score that must be played by either a synthesizer or by
humans to produce an audio music performance. This project focuses on
generative modeling of symbolic (MIDI) music to generate original
musical compositions.

@@

The state of the art in music generation has a long way to go before
it can consistently generate music scores or performances that would
be enjoyable and popular for humans to listen to, but a number of
recent research projects have shown promising progress in this area.

@@

MusicVAE, part of the Magenta project, is a variational Long
Short-Term Memory (LSTM) autoencoder for MIDI that incorporates a
novel hierarchical structure using a ``conductor'' recurrent layer in
its decoder model to better capture structure at multiple levels and
avoid the problem of ``posterior/mode collapse'' whereby a generative
model learns to ignore its latent code and rely on autoregression
\cite{roberts_hierarchical_2018}. This model is capable of generating
new melodies that blend two given melodies via latent space
interpolation.

Another Magenta model called Music Transformer is a generative model
that borrows its approach from the Natural Language Processing (NLP)
domain, using a self-attention network to model MIDI music as a
sequence of discrete tokens with relative positional dependencies
\cite{huang_music_2018}. The focus of this model is on learning
long-term dependencies in music to produce longer clips of music with
coherent structure. Music Transformer was trained on a dataset of
Piano-e-competition performances \cite{hawthorne2019enabling} and its
generated piano music received favorable qualitative (Likert scale)
ratings from human listeners for its resemblance to human-composed
music \cite{huang_music_2018}.

MuseGAN \cite{dong2017musegan} is an application of Generative
Adversarial Networks (GAN) to polyphonic MIDI music generation,
trained on four-bar phrases of a multi-track pianoroll representation
of rock songs from the Lakh Midi Dataset
\cite{raffel_learning-based_2016}. Like MusicVAE, MuseGAN includes a
two-level generator that first samples latent codes at the phrase or
bar level, then generates notes within the bars, to produce
longer-term structural patterns.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{song_structure.png}
  \caption{Diagram of the hierarchical structure of a musical
    composition, from the MuseGAN paper \cite{dong2017musegan}.}
  \label{song_structure}
  \Description{A diagram showing the hierarchical structure of songs,
    to paragraphs, to phrases, to bars, to, beats, to pixels.}
\end{figure}

A major advantage of working with the symbolic representation of music
is that it is of far lower dimensionality than the raw audio waveforms
of a recorded performance, which makes it less computationally
expensive. However, there are many aspects of musical performance
style that are not captured by a symbolic representation, so the
expressiveness of symbolic generative models is limited
\cite{manzelli_conditioning_2018}.

@@

Prior work points out that the division between symbolic music notes
and music performances, is analogous to the division between symbolic
language and utterances in speech, which may inspire ideas for
combining the two approaches \cite{hawthorne2019enabling}. A paper
from Boston University describes an effort to combine the symbolic and
waveform approaches to music modeling, by training an LSTM to learn
melodic structure of different styles of music, then providing
generations from this model as conditioning inputs to a WaveNet-based
raw audio generator \cite{manzelli_conditioning_2018}.

@@

This research project will focus on the generative modeling of
symbolic music using MIDI data as inputs, because of the advantages of
symbolic music models in representing long-range patterns in musical
compositions to produce generations with coherent structure and use of
repetition over long time scales.

@@

Two primary datasets will be used for this research project: MusicNet,
which is a collection of 330 freely licensed European classical music
recordings with aligned MIDI scores \cite{thickstun2017learning}, and
the Lakh MIDI Dataset, which includes a collection of 45,129 Creative
Commons licensed MIDI files of popular music songs that have been
matched and aligned to MP3 recordings, and of which I plan to use a
subset for model training and evaluation
\cite{raffel_learning-based_2016}. Including this second dataset may
help us to generalize the modeling approach beyond European classical
music to include other popular genres and associated instruments.

@@

Several choices must be made in how to preprocess binary MIDI files
into training examples for a neural network. There are multiple
open-source Python packages that assist with the process of reading
MIDI files from their binary on-disk representations into Python
objects, such as pretty\_midi \cite{raffel_pretty_midi_2014} and
Pypianoroll \cite{dong_pypianoroll_2018}.

In order to accommodate polyphonic music, I will convert each MIDI
file into a pianoroll representation, as visualized in Figure
\ref{pianoroll}, wherein each instrument track is a sparse matrix that
multi-hot encodes the velocity values for each of 128 possible pitch
levels at each timestep. To reduce dimensionality, I will clip the
note pitch values to a narrower range, excluding the rarely used notes
in the extreme high and low octaves, and encode chords (combinations
of notes played simultaneously on the same instrument) as distinct
tokens, so that

Because songs are typically at least a few minutes long and of varying
length, it will not be feasible to train with entire songs as
examples, so we will crop songs into phrases of equal numbers of
measures to use as training data.

@@

While I plan to model music with multiple instrument tracks, I
anticipate the need to model only a fixed selection of instrument
parts, similar to how MusicVAE models three-part (drum, bass and
melody) \cite{roberts_hierarchical_2018} and MuseGAN models five-part
(drum, bass, guitar, string, piano) arrangements.

\begin{figure}[h]
  \centering
  \includegraphics[width=\linewidth]{first_bar.png}
  \caption{Pianoroll visualization of the first measure of
    Beethoven's Serioso String Quartet}
  \label{pianoroll}
  \Description{A pianoroll representation of cello, viola and violin
    parts from a Beethoven piece.}
\end{figure}

Data augmentation is also possible and I plan to test its impact on
the results--the literature suggests augmentation via pitch shifting
each training example up or down by up to six semitones, and
increasing or reducing the speed by up to 10\% in order to create
additional training examples and reduce overfitting
\cite{oore_this_2018}.

\subsection{Model Design}

I plan to utilize recurrent neural networks with LSTM or GRU cells to
model the sequential structure of music, stacking multiple layers to
allow the network to learn at multiple levels of abstraction.

There are two general methods of generating sequences from a model:
sliding window prediction, where a model is trained on a fixed length
sequence and then used to predict the next item in the sequence
repeatedly; and latent space interpolation, where a model learns
random distributions of latent code embeddings for the sequences and
draws from the distributions to generate new samples. I plan to
attempt both of these approaches and compare the results.

@@

Evaluation of generative models is challenging because there is no
equivalent of an accuracy metric like what is used in supervised
learning. For autoencoder models we can measure how accurately the
model can reconstruct its own inputs, but this does not tell us the
quality of the interpolated examples. Generative models are typically
evaluated using a combination of qualitative metrics whereby human
judges rate the quality of the generated examples (essentially a
Turing test), and quantitative metrics that assess the differences in
the parametric distributions of generated and real examples. Yang and
Lerch (2020) proposes a set of metrics informed by music theory, for
probabilistically evaluating how similar the generations are to known
sample distributions of real music \cite{yang_evaluation_2020}. These
metrics include counts, ranges, histograms and transition matrices of
pitches and note lengths, then the Kullback-Leibler divergence and
overlapping area of the probability density functions are used to
compare against known reference distributions per musical genre
\cite{yang_evaluation_2020}. Due to the cost and time requirements
associated with designing a human subjects experiment, we plan to
utilize this quantitative approach to the quality assessment of
generated samples.

@@

At this stage, I have decided on MIDI generation as the modeling
problem, downloaded several MIDI datasets, and begun exploring them to
learn how to interpret, process, and visualize them.

I have implemented several pianoroll data processing functions that
crop sections of MIDI files into fixed-length training examples
consisting of multi-track note sequences in symbolic time (beats) at a
given time resolution. I have started implementing a training data
processing pipeline that will read MIDI files from disk and transform
them into the pianoroll-style tensor representation, providing batches
of training examples to Keras model layers for fitting.

I have started designing a simple LSTM autoencoder to model short
(e.g. 16 beats or 4 bars) clips of monophonic (single instrument
track) music, in order to learn the basics of sequence model
architecture. I will create a sequence predictor model with a similar
architecture and compare it to the autoencoder. Next I will add
support for polyphonic music with 3-5 instrument tracks. Once
polyphonic music generation is working well, I plan to increase the
length of the sequences and make adjustments to the model architecture
to permit it to model longer phrases and paragraphs of music.

I aim to build upon the hierarchical decoding approach taken in the
MusicVAE \cite{roberts_hierarchical_2018} and MuseGAN
\cite{dong2017musegan} papers, possibly adding more than two recurrent
layers in the decoder, to address a key challenge in generative music
modeling: the ability to learn the layered patterns in music to
produce compositions with convincing long-term structure.
